The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-03-28 Paper.Rmd
working on Discussion

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
exceedingly common in datasets that range across multiple sub-disciplines
and levels of organization in ecology, in spite of the fact that no published 
ecological research we are aware of has ever
taken this structure into account. Via simple mathematical models, we hypothesized two 
mechanisms that may cause non-N copula structures with asymmetric tail dependence
in ecological data. One
mechanism was via a Moran-like effect by which the dependence structure between ecological
variables, measured through time in different locations, is shaped by the dependence structure of 
an environmental variable in those locations that influences the ecological variable. 
A second mechanism gives rise to non-N copula structure and asymmetric tail dependence via
asymmetric or nonlinear influence of environmental variables on ecological variables, 
and relates to Liebigs law of the minimum. We discuss below how commonly these and other
potential mechanisms may opperate. 
<!--***DAN: make sure to do so-->
We also demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology. First, the skewness, through time, 
of spatially averaged time series is affected by the tail dependence of the relationships
between the local quantities being averaged (such as local abundances of a pest or a resource). 
Thus right-tail-dependent local quantities 
tend to produce "spiky" spatially averaged time series, with large outbreaks. 
Second, metapopulation extinction risk can be accentuated or reduced by the tail dependence
properties of environmental variables that influence the populations. Finally, we presented
simulation results in an Appendix showing that spatial Taylor's law can be strongly affected by
tail dependence between local populations. In our view, our results show that
many ecological phenomena cannot be completely understood without using a more complete, 
copula-based approach to analyzing dependence between variables, and we advocate that
such approaches become standard tools in ecology.

Put ideas here for discussion paragraphs:
1) Maybe a para on how likely one is to get copula structure via the Liebig mechanism - considers
what people have written on nonlinear effects and empirical evidence for Liebig's law and so on
2) Maybe a para on how likely one is to get copula structure via the Moran mechanism - considers
work on copula structure in environmental variables. Future work: copula structure of environmental
variables known to be important in ecology, including how it may be changing and whether it has a 
geography.
3) a para on stability of the total biomass time series, and skewness - sets up Shya's later paper as
future work
4) We have only given two mechanisms, but there are likely to be others. Maybe I can think
of some ideas and make a paragraph.
5) If Van gives us some ideas, they may fit in the Discussion.
6) Maybe give Holder a bunch of draws from a bivariate dist with standard normals marginals but with 
some tail-dependent copula structure and ask him to simulate evolution down a mammal or bird tree and
see what you get. Then give him the other tail dependence, and no tail dependence (bivariate standard
normal). Maybe this will lead to more results, and also a Discussion para on why evolutionary steps 
might be tail dependent. A result that tail dependent steps lead to tail dependent character dists
in the end would replace one question (why are there tail-dependent characters) with another (why are
there tail dependent steps). Of course it could also be the case that steps are not tail dependent and 
the tail dependence of characters arises in another way. Would also lead in to alternative thinking in 
evolutionary biology - Mark could write a para on why it would matter for evolutionary biol if there are 
tail dependent steps.
7) A para that sets up Shya's extinction risk paper - asks about importance of this effect for real 
species and whether it needs to be incorporated into PVA, and in what cases. If the copula structure of environmental variables changes (is it changing?) what will be the impact on extinction risk? Also, there may be copula structure between vital rates, and it is not accounted for in typical stochastic matrix model based PVA. Does it need to be?
8) A para about what is available in the copula and VineCopula packages. Maybe make it a broader para about emthods, and state in there that we developed some new surrogate methods that may be broadly useful for researches trying to incorporate copula approaches.

>>>>

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
exceedingly common in datasets that range across multiple sub-disciplines
and levels of organization in ecology, in spite of the fact that no published 
ecological research we are aware of has 
taken this structure into account. Via simple mathematical models, we hypothesized two 
mechanisms that may cause non-N copula structures with asymmetric tail dependence
in ecological data. One
mechanism was via a Moran-like effect by which the dependence structure between ecological
variables (measured through time at different locations) is shaped by the dependence structure of 
environmental variables (measured at the same times and locations). 
A second mechanism gives rise to non-N copula structure and asymmetric tail dependence via
the asymmetric or nonlinear influence of environmental variables on ecological variables, 
and relates to Liebig's law of the minimum. We discuss below how commonly these and other
potential mechanisms may opperate. 
<!--***DAN: make sure to do so-->
We also demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology. First, the skewness, through time, 
of the spatial average of time series from different locations (e.g., local pest abundance
time series) is affected by the tail dependence of the relationships
between the local quantities. Thus right-tail-dependent local quantities 
tend to produce "spiky" spatially averaged time series, with large outbreaks. 
Second, metapopulation extinction risk can be accentuated or reduced by the tail dependence
properties of environmental variables that influence the populations. Finally, 
simulation results in an Appendix showed that spatial Taylor's law can be strongly influenced by
tail dependence between local populations. In our view, our results show that, for many
ecological phenomena, a complete understanding is impossible without 
characterizing dependence between variables using copulas.

<!--8) A para about what is available in the copula and VineCopula packages. Maybe make it a broader para about emthods, and state in there that we developed some new surrogate methods that may be broadly useful for researches trying to incorporate copula approaches.-->
Copula approaches are
well developed, statistically [@joe2014_dependence; @nelsen2006_copula], and 
convenient computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in R). Ecologists need only to become aware of these tools
and understand them conceptually to start applying them broadly to ecological
problems. We created a variety of interrelated surrogate (randomization) procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that build upon existing methods and may be especially useful in ecological applications.

<!--***DAN: some mechanism paragraphs follow-->

<!--Maybe a para on how likely one is to get copula structure via the Moran mechanism - considers
work on copula structure in environmental variables. Maybe mention possible future work: 
study copula structure of environmental variables (such as temperature and precip) that are known 
to be important in ecology, including how such copula structure may be changing and whether it 
has a geography of the kind studied in Walter et al.-->
<!--***DAN: Shya, please read or reread all the papers we have that say anything about 
copula structure of dependence between an environmental variable measured in different places, 
and then write this paragraph.-->
One may expect our Moran-like mechanism for non-N copula structure and tail dependence to 
operate commonly, because Moran effects are likely ubiquitous [@Liebhold2004], and 
non-N and tail-dependent copula structures are common in environmental variables.
<!--***DAN: Shya, this topic sentence may have to be modified if you find non-N and tail-dependent
copula structures in environmental variables measured in different places are not so common,
or have not been studied much. Butif you find some papers, and they show non-N structure, you 
can just summarize their results here.-->
<!--***DAN: Idea to work in if it fits: if large weather events (e.g., large rain events) 
are also widespread, that should lead to right-tail dependence in, e.g., rainfall.-->

<!--Maybe a para on how likely one is to get copula structure via the Liebig mechanism - considers
what people have written on nonlinear effects and empirical evidence for Liebig's law and so on-->
There are also substantial reasons to expect that our second mechanism operates commonly. 
The second mechanism is the one that involves asymmetric or nonlinear influences of 
environmental variables. Liebig's law of the minimum and the idea of the "limiting 
nutrient" for organism or population growth are fundamental and dominant paradigms in ecology.
Furthermore, innumerable studies document nonlinear influences of environmental
variables on important ecological quantities at all levels of organization, e.g., the 
influence of temperature on: 
individual metabolism and autotroph growth rates 
[@GarciaCarreras2018]; dispersal [@Oconnor2007]; population growth 
[@Savage2004]; and microbial and ecosystem methane flux [@YvonDurocher2014].
<!--***DAN: you may want to combine the above para with the previous one in the end, using an
intro something like "There are good reasons to suspect that the two mechanisms we suggested
operate extremely commonly in real ecosystems"-->

<!--4) We have only given two mechanisms, but there are likely to be others, and discoverig
additional mechanisms of copula structure is certain worthy of future work. Maybe I can think
of some ideas and make a paragraph.
5) If Van gives us some ideas, they may fit in the Discussion here.
6) Maybe give Holder a bunch of draws from a bivariate dist with standard normals marginals but with 
some tail-dependent copula structure and ask him to simulate evolution down a mammal or bird tree and
see what you get. Then give him the other tail dependence, and no tail dependence (bivariate standard
normal). Maybe this will lead to more results, and also a Discussion para on why evolutionary steps 
might be tail dependent. A result that tail dependent steps lead to tail dependent character dists
in the end would replace one question (why are there tail-dependent characters) with another (why are
there tail dependent steps). Of course it could also be the case that steps are not tail dependent and 
the tail dependence of characters arises in another way. Would also lead in to alternative thinking in 
evolutionary biology - Mark could write a para on why it would matter for evolutionary biol if there are 
tail dependent steps.
-->

<!--***DAN: some consequences paragraphs next-->

<!--a para on stability of the total biomass time series in a community, and skewness - sets up Shya's later paper as future work-->

<!--A para that sets up Shya's extinction risk paper - asks about importance of this effect for real 
species and whether it needs to be incorporated into PVA, and in what cases. If the copula structure of environmental variables changes (is it changing?) what will be the impact on extinction risk? Also, there may be copula structure between vital rates, and it is not accounted for in typical stochastic matrix model based PVA. Does it need to be?-->
Although we have demonstrated that tail dependence between environmental variables
can influence extinction risk, substantial work remains to be done regarding the 
circumstances under which this result holds, and its magnitude in different scenarios. 
First, we used 
a non-density dependent model to examine extinction risk; does tail dependence
in environmental variables still influence extinction risk when density dependence
plays a role? Second, we considered metapopulation extinction risk, but the whole field 
of population viability analysis [@REF-Morris and Doak] stochastic matrix modelling

<!--Maybe a para from Holder on his thoughts in the importance of copulas for
macroevolutionary biology-->

<!--Conclusion of some kind-->
